---
title: "Milestone 2"
names: "Dawson Coblin & Jose Lazo"
date: "October 3, 2022"
output: 
---
```{r, echo=FALSE}
library(readxl)
library(tidyr)
library(lubridate)

```

```{r}
MPX_data <- read.csv("~/251/files/euro_mpx_cases.csv")
```

Description of dataset

o	What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
The monkeypox data file contains information on the number of monkeypox cases reported by EU/EEA countries or collected throughout epidemiologic intelligence at ECDC. 
Each row contains the corresponding data for a country, day of reporting, number of cases and source of information (data are in long format). The file is updated twice a week. The date range for this data set is from 05/09/22 - 08/23/22

o	How does the dataset relate to the group problem statement and question?
For our project we are interested in understanding how case rates may differ by region and various demographic factors. The MPX data has information on confirmed cases by countries.

/newpage

•	Import statement
o	NOTE: Please use datasets available in the PHW251 Project Data github repo Links to an external site. (this is important to make sure everyone is using the same datasets)
o	Use appropriate import function and package based on the type of file
o	Utilize function arguments to control relevant components (i.e. change column types, column names, missing values, etc.)
o	Document the import process

Import process
1) Download zip file from phw251_projectdata
2) Set up our github repository on our R datahub
3) One team member uploaded pertinent excel files to their Rproject 251
4) Commit and push to have excel files on our team's github

/newpage

•	Identify data types for 5+ data elements/columns/variables
o	Identify 5+ data elements required for your specified scenario. If <5 elements are required to complete the analysis, please choose additional variables of interest in the data set to explore in this milestone.
o	Utilize functions or resources in RStudio to determine the types of each data element (i.e. character, numeric, factor)
o	Identify the desired type/format for each variable—will you need to convert any columns to numeric or another type? 

```{r}
str(MPX_data)
mean(MPX_data$ConfCases)
```
/newpage

•	Provide a basic description of the 5+ data elements
o	Numeric: mean, median, range
o	Character:  unique values/categories
o	Or any other descriptives that will be useful to the analysis

Confcases is a numeric data element
```{r}
round(mean(MPX_data$ConfCases),2)
median(MPX_data$ConfCases)
range(MPX_data$ConfCases)
```
Mean = 5.72
Median = 0
Range = 0,655

DateRep 
Is a character variable which contains dates 05/09/22 through 08/23/22. We will convert this data into numeric form.

```{r}
unique(MPX_data$CountryExp)
```

CountryExp
Is a character variable which contains 29 countries listed above.

```{r}
unique(MPX_data$CountryCode)
```

CountryCode
Is a character with the 2 letter abbreviation of countries with 29 values which matches with CountryExp


```{r}
unique(MPX_data$Source)
```

Source is a character variable with 2 options, TESSy and EI
